---
title: "DATS 6101-10 Project 2: Predicting the Number of Abalone Growth Rings using a Blacklip Abalone Dataset"
author: "Group Members: Jessica Barahona, Jesse Liu, and Solomon Mekonnen"
date: "Presented on April 22nd, 2020"
output:
  html_document:
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cerulean 
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = 'markup', message = F)
knitr::opts_chunk$set(warning = F, results = 'hide', message = F)
# knitr::opts_chunk$set(include = F)
# knitr::opts_chunk$set(echo = TRUE)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r basic, include=F}
# use this function to conveniently load libraries and work smoothly with knitting
# can add quietly=T option to the require() function
loadPkg = function(pkg, character.only = FALSE) { 
  if (!character.only) { pkg <- as.character(substitute(pkg)) }
  pkg <- ifelse(!character.only, as.character(substitute(pkg)) , pkg)  
  if (!require(pkg,character.only=T, quietly =T)) {  install.packages(substitute(pkg),dep=T); if(!require(pkg,character.only=T)) stop("Package not found") } 
}
loadPkg(knitr)

# unload/detact package when done using it
unloadPkg = function(pkg, character.only = FALSE) { 
  if(!character.only) { pkg <- as.character(substitute(pkg)) } 
  search_item <- paste("package", pkg,sep = ":") 
  while(search_item %in% search()) { detach(search_item, unload = TRUE, character.only = TRUE) } 
}
```

```{r xkablesummary, include=F}
loadPkg(xtable)
loadPkg(kableExtra)
loadPkg(stringi)

xkabledply = function(smmry, title='Caption', pos='left') { # Thanks Ryan Longmuir for the codes
  smmry %>%
    xtable() %>% 
    kable(caption = title, digits = 4) %>%
    kable_styling(position = "center") %>%
    kable_styling(bootstrap_options = "striped", full_width = F,
    position = pos)
}

xkablesummary = function(df) { 
  #' Combining base::summary, xtable, and kableExtra, to easily display numeric variable summary of dataframes. 
  #` If the categorical variables has less than 6 levels, the function will still run without error.
  #' ELo 202003 GWU DATS
  #' version 1
  #' @param df The dataframe.
  #' @return The summary table for display, or for knitr to process into other formats 
  #' @examples
  #' xkablesummary( faraway::ozone )
  #' xkablesummary( ISLR::Hitters )
  
  s = summary(df) %>%
    apply( 2, function(x) stringr::str_remove_all(x,c("Min.\\s*:\\s*","1st Qu.\\s*:\\s*","Median\\s*:\\s*","Mean\\s*:\\s*","3rd Qu.\\s*:\\s*","Max.\\s*:\\s*")) ) %>% # replace all leading words
    apply( 2, function(x) stringr::str_trim(x, "right")) # trim trailing spaces left
  
  colnames(s) <- stringr::str_trim(colnames(s))
  
  if ( dim(s)[1] ==6 ) { rownames(s) <- c('Min','Q1','Median','Mean','Q3','Max') 
  } else if ( dim(s)[1] ==7 ) { rownames(s) <- c('Min','Q1','Median','Mean','Q3','Max','NA') }
  
  s %>%
    xkabledply("Table: Statistics summary.", "center")

}

xkablevif = function(model) { 
  #' Combining faraway::vif, xtable, and kableExtra, to easily display numeric summary of VIFs for a model. 
  #' ELo 202003 GWU DATS
  #' version 1
  #' @param df The dataframe.
  #' @return The summary table for display, or for knitr to process into other formats 
  #' @examples
  #' xkablevif( model )
  
  vifs = table( names(model$coefficients)[2:length(model$coefficients)] ) # remove intercept to set column names
  vifs[] = faraway::vif(model) # set the values

  vifs %>%
    xtable() %>% 
    kable(caption = "VIFs of the model", digits = 4, col.names = 'VIF') %>% # otherwise it will only has the generic name as 'V1' for the first vector in the table
    kable_styling(position = "center") %>%
    kable_styling(bootstrap_options = "striped", full_width = F,
    position = "left")
}
```

```{r importing, echo = FALSE}

#importing data
abalone_data <- data.frame(read.csv("abalone.csv"))

#importing packages 
loadPkg("ggplot2")
loadPkg("dplyr")
loadPkg("corrplot")
loadPkg("stats")
loadPkg("modelr")
loadPkg("stringr")
loadPkg("faraway")
loadPkg("tidyr")
loadPkg("gridExtra")
```

# Project Introduction:

Abalone are marine snails belonging to the class Gastropoda, family Haliotidae, and genus Haliotis. Abalone are found in cold waters around most parts of the world, including the northwestern coast of the United States and Canada as well as South Africa and New Zealand. Abalone is highly sought out for its meat and is considered one of the most expensive shellfish in the world. Abalone can cost as high as 500 USD per kilo, which equates to approximately 250 grams of meat. Additionally, the abalone shell has an inner, iridescent layer that is used in decorative pieces and jewelry. Unfortunately, there has been a steep decline in the abalone wild population over the years due to overfishing and environmental stressors. In recent years, several countries have placed restrictions on the total number of abalone, age of abalone, and/or size of abalone that can be collected per person to combat declining populations. For example, in South Africa, the minimum legal size is 114 mm shell breadth, which is considered around 8 to 9 years of age. With this in mind, we found a dataset from the UCI Machine Learning Repository that contains 4,177 observations across 9 variables for a population of blacklip abalone. One of the most difficult tasks is estimating the age of an abalone, which traditionally involves staining the shell and counting the number of growth rings individually under a microscope. In our analysis, we will analyze the relationships between the variables to then build a model to predict the number of growth rings on an abalone. Our analysis and model may lay the groundwork for future methods enlisted to develop policies and/or enforce restrictions on abalone fishing. 

# Background Research on the Abalone Dataset:

The dataset is from the UCI Machine Learning Repository. The original data is from a population study of the blacklip abalone carried out in Tasmania. 

Below is detailed information regarding the original study:

*Original Study*

Nash, W.J., Sellers, T.L., Talbot, S.R., Cawthorn, A.J., & Ford, W.B. (1994). The Population Biology of Abalone (Haliotis species) in Tasmania. I. Blacklip Abalone (H. rubra) from the North Coast and Islands of Bass Strait. Technical Report No. 48, Sea Fisheries Division, pp. 1-69.

*Original Owners of Database* 

Marine Resources Division 
Marine Research Laboratories - Taroona 
Department of Primary Industry and Fisheries, Tasmania 
GPO Box 619F, Hobart, Tasmania 7001, Australia 

*Donor of Database*

Sam Waugh (Sam.Waugh '@' cs.utas.edu.au) 
Department of Computer Science, University of Tasmania 
GPO Box 252C, Hobart, Tasmania 7001, Australia 

From the original data, examples with missing values were removed (the majority having the predicted value missing), and the ranges of the continuous values have been scaled for use with an ANN (by dividing by 200).

2.2 Limitations of the Dataset and Considerations

The dataset is limited to an abalone population in Tasmania and captures measurements for the blacklip abalone (Haliotis rubra). Growth checks and layers in the shell were used to age the abalone assuming the one-layer-per-year relationship established for blacklip abalone in southern Tasmania. The addition of 1.5 years to the number of major growth rings was based on a finding from Prince et al. Three distinct minor layers are laid down in the first 16 months of life, with subsequent major layers being deposited annually. The 1.5 years accounts for the period during which the minor rings are deposited. The validity of this application to backlip abalone populations beyond southeast Tasmania is unknown (Nash et al, 1994, pg. 9).

This well studied relationship of age = number of major growth rings + 1.5 years is unique to blacklip abalone. The relationship between the number of major growth rings and age for other species of abalone is unknown or unclear. Therefore, we chose to keep number of major growth rings as our y-target variable and not convert the "Rings" variable to age for the purposes of our analysis. 

# S.M.A.R.T Question

Can we predict the number of growth rings on an abalone from a specific set of variables?

# Exploratory Data Analysis: Analyzing the Abalone Dataset

The dataset contains 4,177 observations across 9 variables. Each row corresponds to a single abalone. The 9 variables are sex, length, diameter, height, whole weight, shucked weight, viscera weight, shell weight, and rings. All variables, except for sex, are numerical.  

 - Sex: Male (M), Female (F), and Infant (I)
 - Length: longest shell measurement (mm)
 - Diameter: perpendicular to length (mm)
 - Height: with meat in shell (mm)
 - Whole weight: whole abalone (g)
 - Shucked weight: weight of meat (g)
 - Viscera weight: gut weight (after bleeding) (g)
 - Shell weight: after being dried (g)
 - Rings: number of major growth rings present


```{r summary}
# Statistics summary
xkablesummary(abalone_data)
#nrow(abalone_data1)
```
 
# Exploratory Data Analysis

```{r normalitytests, echo=FALSE}
loadPkg(ggplot2)

#histogram distribution for Rings
ggplot(abalone_data, aes(x=Rings, color=Sex))+
  geom_density(aes(fill=Sex), alpha=0.4)+
   labs(title = "Distribution of Abalone Rings",
       x = "Rings",
       y = "Frequency")

#histogram distribution for Length
ggplot(abalone_data, aes(x=Length, color=Sex))+
  geom_density(aes(fill=Sex), alpha=0.4)+
   labs(title = "Distribution of Abalone Length",
       x = "Length (mm)",
       y = "Frequency")

#histogram distribution for Diameter
ggplot(abalone_data, aes(x=Diameter, color=Sex))+
  geom_density(aes(fill=Sex), alpha=0.4)+
   labs(title = "Distribution of Abalone Diameter",
       x = "Diameter (mm)",
       y = "Frequency")

#histogram distribution for Height
ggplot(abalone_data, aes(x=Height, color=Sex))+
  geom_density(aes(fill=Sex), alpha=0.4)+
   labs(title = "Distribution of Abalone Height",
       x = "Height (mm)",
       y = "Frequency")

#histogram distribution for weight
ggplot(abalone_data, aes(x=Whole.weight, color=Sex))+
  geom_density(aes(fill=Sex), alpha=0.4)+
   labs(title = "Distribution of Whole Weight",
       x = "Whole Weight (grams)",
       y = "Frequency")

#histogram distribution for Shuchked Weight
ggplot(abalone_data, aes(x=Shucked.weight, color=Sex))+
  geom_density(aes(fill=Sex), alpha=0.4)+
   labs(title = "Distribution of Shucked weight",
       x = "Shucked weight (grams)",
       y = "Frequency")

#histogram distribution for Viscera Weight
ggplot(abalone_data, aes(x=Viscera.weight, color=Sex))+
  geom_density(aes(fill=Sex), alpha=0.4)+
   labs(title = "Distribution of Viscera Weight",
       x = "Viscera Weight (grams)",
       y = "Frequency")

#histogram distribution for Shell Weight
ggplot(abalone_data, aes(x=Shell.weight, color=Sex))+
  geom_density(aes(fill=Sex), alpha=0.4)+
   labs(title = "Distribution of Shell Weight",
       x = "Shell Weight (grams)",
       y = "Frequency")
```

```{r qqtests echo=FALSE}

qqnorm(abalone_data$Rings) 
qqline(abalone_data$Rings)
#Shows a right skew

qqnorm(abalone_data$Length) 
qqline(abalone_data$Length)
#Shows a left tailed distribution

qqnorm(abalone_data$Diameter) 
qqline(abalone_data$Diameter)
#Shows a left tailed distribution

qqnorm(abalone_data$Height) 
qqline(abalone_data$Height)
#normal distribution

qqnorm(abalone_data$Whole.weight) 
qqline(abalone_data$Whole.weight)
#Shows a right tailed distribution

qqnorm(abalone_data$Viscera.weight) 
qqline(abalone_data$Viscera.weight)
#rightskew

qqnorm(abalone_data$Shucked.weight) 
qqline(abalone_data$Shucked.weight)
#rightskew

qqnorm(abalone_data$Shell.weight) 
qqline(abalone_data$Shell.weight)
#rightskew
```

```{r agescat, echo=FALSE}

ggplot(abalone_data, aes(x=Rings, y=Length)) + 
  geom_point(aes(col=Sex)) + 
  labs(subtitle="Abalone Rings vs. Length", 
       y="Length", 
       x="Rings", 
       title="Scatterplot")

ggplot(abalone_data, aes(x=Rings, y=Diameter)) + 
  geom_point(aes(col=Sex)) + 
  labs(subtitle="Abalone Rings vs. Diameter", 
       y="Diameter(mm)", 
       x="Rings", 
       title="Scatterplot")

ggplot(abalone_data, aes(x=Rings, y=Height)) + 
  geom_point(aes(col=Sex)) + 
  labs(subtitle="Abalone Rings vs. Height", 
       y="Height(mm)", 
       x="Rings", 
       title="Scatterplot")

ggplot(abalone_data, aes(x=Rings, y=Whole.weight)) + 
  geom_point(aes(col=Sex)) + 
  labs(subtitle="Abalone Rings vs. Whole Weight", 
       y="Whole Weight (grams)", 
       x="Rings", 
       title="Scatterplot")

ggplot(abalone_data, aes(x=Rings, y=Shucked.weight)) + 
  geom_point(aes(col=Sex)) + 
  labs(subtitle="Abalone Rings vs. Shucked Weight (grams)", 
       y="Shucked Weight (grams)", 
       x="Rings", 
       title="Scatterplot")

ggplot(abalone_data, aes(x=Rings, y=Viscera.weight)) + 
  geom_point(aes(col=Sex)) + 
  labs(subtitle="Abalone Rings vs. Viscera Weight (grams)", 
       y="Viscera Weight (grams)", 
       x="Rings", 
       title="Scatterplot")

ggplot(abalone_data, aes(x=Rings, y=Shell.weight)) + 
  geom_point(aes(col=Sex)) + 
  labs(subtitle="Abalone Rings vs. Shell Weight (grams)", 
       y="Shell Weight (grams)", 
       x="Rings", 
       title="Scatterplot")
```


```{r ageboxplot, echo=FALSE}
loadPkg("dplyr")

abalone %>% group_by(Rings) %>% boxplot(abalone$Length,abalone$Rings)

```

### Bivariate Analysis on quantitative features

```{r cor}
loadPkg(psych) # pair plots with histogram on diagonal and other options [,2:9]
pairs.panels(abalone_data, 
             method = "pearson", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             ellipses = FALSE # show correlation ellipses
             )
#unloadPkg(psych)
```
### Correlation Analysis
 
## Analysis on the categorical variable - Sex

```{r Cat_var_sex, echo=FALSE}
ggplot(abalone_data, aes(x= Sex, y=Rings, fill=Sex))+
  geom_boxplot(alpha=0.3)+
  labs(x="SEX", y="Number of rings", caption = "", title = "Number of Rings Distribution among Male, Female and Infant Abalone")
```
 - The number of Rings difference between Males and Females seem to be insignificant. The I category seems to show a significant difference but this is not due to a function of Sex category rather the function of Age i.e. younger Abalones have fewer rings. 

## Model Selection

### Multiple Linear Regression Model


```{r lm1, echo=True}
ml_model <- lm(Rings~., data = abalone_data)
summary(ml_model)
vif(ml_model)
```
- The Variable Inflation Factors(VIF) for the model indicates multicollinearity between all factors, except for Male (1.40), Infant(1.98) and Height (3.58), which have less values (<5). The high VIF values suggests that there are redundant predictors. These variables are correlated, the standard OLS parameter estimates makes unreliable model.

```{r}
loadPkg(ggfortify)
autoplot(ml_model, label.size=3, which=c(2,0)) #QQ plot
```
-This plot shows that the model residuals are not normally distributed - indicating that the linear model is not a good fit.

- This figure shows that the model residuals are not normally distributed - indicating there are higher order effects that the linear model is not capturing. 

### Principal components regression

```{r}
loadPkg(dummies)
abalone_data <- dummy.data.frame(abalone_data,"Sex")
pca =prcomp(abalone_data, scale =TRUE)
pca$rotation
summary(pca)

```


```{r PCR}
loadPkg(pls)
set.seed (1000)
 
pcr_model <- pcr(Rings~., data = abalone_data, scale = TRUE, validation = "CV")
 
summary(pcr_model) 

```
```{r}
#Mean square error rate
validationplot(pcr_model,val.type = "MSEP")
```

```{r, results='show', include=T}
#pcr_model$coefficients[1:8,1,'1 comps'] # only one coefficient for PC1, but expressed in the original variables coefficients.
pcr_model$coefficients[1:8,1,'4 comps'] # three coefficients for PC1, PC2, and PC3, but expressed in the original variables coefficients.
# 
#pcr_model$fitted.values[1:5,1,'1 comps'] # the fitted values. Showing only the first five here.
#pcr_model$fitted.values[1:5,1,'3 comps']
```
 
 

- From the plot we achieved the minimum error rate at 4.Rather than using the orginal values with 9 features , we can reduce the dimenssion to 4 PC.  
Reducing error rate and with smaller number of Principal components.

#kNN Classification: Blacklip Abalone Age Groupings

```{r KNN, echo=FALSE}
#Remove categorial variable - sex variable
abaloneknn <- subset(abalone_data, select =-c(1:3))

#Convert numerical "number of rings" variable to categorical
#Group 1: 1-5 growth rings
#Group 2: 6-13 growth rings
#Group 3: 14-30 growth rings 
abaloneknn$Rings <- cut(abaloneknn$Rings,
                  breaks = c(0,5,13,30),
                  labels = c("Group 1", "Group 2", "Group 3"))

xkablesummary(abaloneknn)
#renaming rings variable to Ring_Group
##abaloneknn <- rename(abaloneknn, Ring_Group = Rings)
##str(abaloneknn)

```

Based on how difficult it is to estimate the age of an abalone, would it be beneficial and easier to estimate which age group an abalone belongs to? We can use the validated relationship between age and number of major growth rings in blacklip abalone to create "number of rings" groups, which correspond to age groups. We can then build our kNN classifier which predicts which group an abalone belongs to. If this kNN classifier predicts with a high accuracy, other kNN classifiers can be built for other abalone species. Researchers, divers, abalone farmers, or fisherman could have a better chance at accurately predicting which age group an abalone belongs to intead of estimating the age. 

Three "number of rings" groups were created based on the life cycle of blacklip abalone and the age assumption (age = number of major growth rings + 1.5 years). "Group 1" is 1-5 major growth rings (<6.5 years old - juvenile), "Group 2" is 6-13 major growth rings (7.5-14.5 years old - adult), and "Group 3" is 14-30 growth rings (15.5-31.5 years old - old adult). 

```{r KNN, results=TRUE}
#ready for knn classification
#first we want to scale the data so KNN will operate correctly
scaledabaloneknn <- as.data.frame(scale(abaloneknn[1:7], center = TRUE, scale = TRUE))
set.seed(1000)
#3:1 train-test split 
abaloneknn_sample <- sample(2, nrow(scaledabaloneknn), replace=TRUE, prob=c(0.75, 0.25))
#We then just need to use the new variable to create the test/train outputs, selecting the first four rows as they are the numeric data in the iris data set and we want to predict Species 
abaloneknn_training <- scaledabaloneknn[abaloneknn_sample==1, 1:7]
abaloneknn_test <- scaledabaloneknn[abaloneknn_sample==2, 1:7]

#create y-variables
abaloneknn.trainLabels <- abaloneknn[abaloneknn_sample==1, 8]
abaloneknn.testLabels <- abaloneknn[abaloneknn_sample==2, 8]
```

```{r best K value, results=TRUE}
i=1                          # declaration to initiate for loop
k.optm=1                     # declaration to initiate for loop
for (i in 1:28){ 
    knn.mod <-  knn(train=abaloneknn_training, test=abaloneknn_test, cl=abaloneknn.trainLabels, k=i)
    k.optm[i] <- 100 * sum(abaloneknn.testLabels == knn.mod)/NROW(abaloneknn.testLabels)
    k=i  
    cat(k,'=',k.optm[i],'\n')       # to print % accuracy 
}
```
```{r, results=TRUE}
plot(k.optm, type="b", xlab="K- Value",ylab="Accuracy level")  # to plot % accuracy wrt to k-value
```

List of accuracy rates for k-values up to 28. We see that a k-value of 15 gives us an accuracy rate of 87.7%. On the graph, we can visually see the accuracy rate level off after it peaks at k-value = 15. 

```{r KNN, results=TRUE}
#load package
loadPkg(FNN)
#deploy model
abalone_pred <- knn(train = abaloneknn_training, test = abaloneknn_test, cl=abaloneknn.trainLabels, k=15)
abalone_pred
#install.packages("gmodels")
loadPkg(gmodels)
IRISPREDCross <- CrossTable(abaloneknn.testLabels, abalone_pred, prop.chisq = FALSE)
```
This kNN classifier raises some concerns even though the accuracy rate is 87.7%. Based on the cross table, the kNN classifer predicted "Group 1" when it should have been "Group 2" 6 times and predicted "Group 3" when it should have been "Group 2" 15 times. Group 1 represents the "juvenile" age group of abalone. For conservational purposes, the "juvenile" abalone group is the group that would need to be most protected. Pulling out juvenile abalone before they are sexually mature would set back attempts of replenishing wild abalone populations. We would be satisfied if the accuracy rate were higher and if there were a reduction in the number of false positives. 

#Conclusion

